---
title: "Assignment Week 4"
output: pdf_document
---

### Name: Ioannis Philippides

### ULCN: s1579495

```{r label = preps, echo=FALSE}
rm(list=ls(all=T)) #clears your workspace
```



This assignment consists of two exercises both directed at computing (or approximating) the value
of the integral 
$$I=\frac 1{\sqrt{2 \pi}}\int_1^\infty x^2 e^{-x^2}\,dx.$$



# Exercise 1. Importance sampling

*(The first part is exercise 5.14 from Rizzo's book)* 

- Obtain a Monte Carlo estimate of $I$ using importance sampling. (There are multiple solutions;
make sure that your method is much more efficient than straightforward simulation from a normal distribution
with mean zero.)

```{r}
f <- function(x) { (1/sqrt(2*pi))*(x^2*exp(-x^2))*(x>1)} # set the function
# create empty vector of length 4 for estimation and standard error
SE <- est <- numeric(4) 
M <- 10000 # 10000 samples per distribution
set.seed(1579495)

# use of Normal(1,1) to estimate g(x)
x <- rnorm(M, mean=1, sd=1)
g <- f(x) / dnorm(x, mean=1, sd=1)
est[1] <- mean(g)
SE[1] <- sd(g)

# use of exponential to estimate g(x):
x <- rexp(M) # default rate equals 1
g <- f(x) / dexp(x)
est[2] <- mean(g)
SE[2] <- sd(g)

#  use of gamma distribution:
x <- rgamma(M, shape=2, rate=1)
g <- f(x) / dgamma(x, shape=2, rate=0.9)
est[3] <- mean(g)
SE[3] <- sd(g)

# use of chisquare:
x <- rchisq(M, df=0.7) 
g <- f(x) / dchisq(x, df=0.7)
est[4] <- mean(g)
SE[4] <- sd(g)



```

\newpage

distribution | Normal   | exponential | gamma    | chisquare
-------------|----------|-------------|----------|------------            
   se        | `r SE[1]`|  `r SE[2]`  |`r SE[3]` | `r SE[4]`
-------------|----------|-------------|----------|------------
estimate     |`r est[1]`| `r est[2]`  |`r est[3]`| `r est[4]`
-------------|----------|-------------|----------|------------

We observe that chisquare distribution doesn't perform so well, comparing to the rest distributions. Normal distribution with mean and variance equal 1, has the smallest standard error.



# Exercise 2. Numerical integration



-  Write a function `trapeziumrule` that takes arguments `f, lower, upper, h`,
equal to a function, lower and upper limits of the integral and a step size,
that returns the integral of the function over the interval [`lower, upper`].

```{r}
# function from integration.R file
trapeziumrule <- function(f, lower, upper, h) {
  x <- seq(lower, upper, h)
  y <- f(x)
  
  # the first and last one count only half
  total <- sum(y) - 0.5*y[1] - 0.5*y[length(y)]
  
  return(h*total)
}

```


- Use this function to find an approximation to the given integral $I$.
Try the 4 step sizes $h=0.01, 0.001, 0.0001, 0.0001$ and use an appropriate
value for `upper`.

```{r}
f <- function(x) { (1/sqrt(2*pi))*(x^2*exp(-x^2)) } # set the function
h <- c(0.01, 0.001, 0.0001, 0.00001) # create 4 steps for given h values
I <- numeric(4)
for (i in 1:4){
# As value for upper, 10 would be enough, because the (decreasing) 
# function will be approximately zero
  I[i] <- trapeziumrule(f, lower = 1, upper = 10, h[i])
}
```

""   |  0.01  |  0.001 | 0.0001 | 0.00001|
-----|--------|--------|--------|--------|
 I   |`r I[1]`|`r I[2]`|`r I[3]`|`r I[4]`|
-----|--------|--------|--------|--------| 

\newpage


- Compute an approximation of $I$ by using the  `R` function `integrate`. Read
the help page of the function to see how to handle the upper, infinite limit!

```{r}
integrate(f, lower=1, upper=Inf) 
# so we see that numerical integration, applying trapezium rule, 
# is a pretty accurate method

```



